---
title: "Loan Data Analysis — A Comprehensive Predictive Modeling Study"
author: "Data Science & Linear Models Project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
```

------------------------------------------------------------------------

# Introduction and Objectives

## Project Overview

This study uses a loan applicant dataset to demonstrate **end-to-end predictive modeling** across two major tasks:

| Task | Goal | Target Variable |
|-----------------|-------------------------|---------------------------------------|
| **Regression** | Predict applicant income | `person_income` |
| **Classification** | Predict loan default | `loan_status` (0 = Default, 1 = Repaid) |

Along the way we performed **feature engineering**, **checking OLS assumptions** apply **transformations**, and conducting a **credit-risk analysis** linking income, credit score and default behavior.

## Use Cases Addressed

1.  **Income Prediction (Regression):** Train Linear Regression, Decision Tree, and Random Forest regressors; select the best model on test-set metrics.
2.  **Assumption Checking & Transformation:** Verify linearity, normality, homoscedasticity, and multicollinearity for the linear model; apply log and sqrt transformations where needed.
3.  **Loan Default Prediction (Classification):** Build Logistic Regression, Random Forest, and Decision Tree classifiers; evaluate with AUC-ROC due to class imbalance.
4.  **Credit Risk Analysis:** Explore how income, credit score, and prior defaults interact to drive loan outcomes.
5.  **Feature Engineering:** Create domain-informed variables (loan-to-credit ratio, high-risk flag, age-experience ratio, income brackets) to boost predictive power.

------------------------------------------------------------------------

# Environment Setup

```{r load-libraries}
libs <- c("tidyverse", "dplyr", "gridExtra", "ggplot2", "ggfortify", "broom",
          "ggcorrplot", "corrr", "corrplot", "MASS", "datarium", "randomForest",
          "rpart", "Hmisc", "caret", "Metrics", "car", "rpart.plot", "e1071", "pROC")

for (lib in libs) {
  if (!(lib %in% installed.packages())) install.packages(lib, quietly = TRUE)
  library(lib, character.only = TRUE)
}

# Resolve namespace conflicts
select    <- dplyr::select
mutate    <- dplyr::mutate
filter    <- dplyr::filter
summarise <- dplyr::summarise

theme_set(theme_classic())
```

## Load Dataset

> **Note:** Update the file path below to match your local environment.

```{r load-data}
df <- read_csv("C:/Users/Light House/Desktop/MastersDSA/semThree-DSA/Linear Models/Data_Project/loan_data.csv", show_col_types = FALSE)
head(df)
```

------------------------------------------------------------------------

# Data Understanding

## Structure & Summary Statistics

```{r structure}
str(df)
summary(df)
```

**Key observations from the summary:**

-   `person_age` peaks at 144 — biologically implausible; outlier treatment required.
-   `person_income` spans \$8 000 to \>\$7.2 million — extreme right skew.
-   `person_emp_exp` reaches 125 years — another impossible value.
-   `loan_status` mean ≈ 0.22 → **78 % of loans defaulted** (severe class imbalance).

## Missing Values & Duplicates

```{r missing-and-duplicates}
missing_counts <- colSums(is.na(df))
cat("Missing values per column:\n"); print(missing_counts)
cat("\nTotal missing values:", sum(missing_counts), "\n")

n_dup <- sum(duplicated(df))
cat("Duplicate rows:", n_dup, "\n")
if (n_dup > 0) {
  df <- df %>% distinct()
  cat("After deduplication:", nrow(df), "rows remain.\n")
}
```

**Result:** No missing values and no duplicates — but implausible outliers still need handling.

------------------------------------------------------------------------

# Exploratory Data Analysis (EDA)

## Univariate Analysis Numeric Variables

```{r univariate-numeric, fig.height=20, fig.width=16}
numeric_vars <- c("person_age", "person_income", "person_emp_exp", "loan_amnt",
                  "loan_int_rate", "loan_percent_income",
                  "cb_person_cred_hist_length", "credit_score")

hist_plots <- lapply(numeric_vars, function(var) {
  ggplot(df, aes(x = .data[[var]])) +
    geom_histogram(bins = 40, fill = "steelblue", color = "white", alpha = 0.8) +
    labs(title = var, x = NULL, y = "Count") +
    scale_x_continuous(labels = scales::comma) +
    scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
    theme(plot.title = element_text(size = 14, face = "bold"),
          axis.text.x = element_text(size = 10, angle = 45, hjust = 1))
})

do.call(grid.arrange, c(hist_plots, ncol = 2))
```

```{r boxplots-numeric, fig.height=14, fig.width=12}
box_plots <- lapply(numeric_vars, function(var) {
  ggplot(df, aes(y = .data[[var]])) +
    geom_boxplot(fill = "steelblue", alpha = 0.6, outlier.color = "red", outlier.size = 1) +
    labs(title = var, y = NULL) +
    scale_y_continuous(labels = scales::comma) +
    theme(plot.title = element_text(size = 11, face = "bold"),
          axis.text.x = element_blank(), axis.ticks.x = element_blank())
})

do.call(grid.arrange, c(box_plots, ncol = 4))
```

**Observations:**

-   `person_age` is right-skewed with extreme outliers above 80.
-   `person_income` is heavily right-skewed a **log transformation** will be essential.
-   `person_emp_exp` contains implausible values (e.g., 125 years).
-   `credit_score` is roughly normally distributed, centered around 630.

### Skewness & Kurtosis

```{r skewness-table}
skew_table <- data.frame(
  Variable = numeric_vars,
  Skewness = sapply(numeric_vars, function(v) round(skewness(df[[v]], na.rm = TRUE), 3)),
  Kurtosis = sapply(numeric_vars, function(v) round(kurtosis(df[[v]], na.rm = TRUE), 3))
)
print(skew_table)
```

> **Rule of thumb:** Skewness beyond ±1 signals the need for transformation in linear modeling.

## Univariate Analysis:Categorical Variables

```{r univariate-categorical, fig.height=10, fig.width=12}
cat_vars <- c("person_gender", "person_education", "person_home_ownership",
              "loan_intent", "previous_loan_defaults_on_file")

cat_plots <- lapply(cat_vars, function(var) {
  ggplot(df, aes(x = reorder(.data[[var]], .data[[var]], function(x) -length(x)))) +
    geom_bar(fill = "steelblue", alpha = 0.8) +
    geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5, size = 3) +
    labs(title = var, x = var, y = "Count") +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          plot.title = element_text(size = 10, face = "bold"))
})

do.call(grid.arrange, c(cat_plots, ncol = 3))
```

## Target Variable Distribution

```{r target-distribution, fig.width=8, fig.height=6}
ggplot(df, aes(x = factor(loan_status), fill = factor(loan_status))) +
  geom_bar(alpha = 0.8, width = 0.5) +
  geom_text(stat = "count",
            aes(label = paste0(after_stat(count),
                " (", round(after_stat(count) / nrow(df) * 100, 1), "%)")),
            vjust = -0.5, size = 5, fontface = "bold") +
  scale_fill_manual(values = c("0" = "#e74c3c", "1" = "#27ae60"),
                    labels = c("Default (0)", "Repaid (1)")) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.12))) +
  scale_x_discrete(labels = c("0" = "Default (0)", "1" = "Repaid (1)")) +
  labs(title = "Loan Status Distribution (Target Variable)",
       subtitle = "Significant class imbalance: ~78% defaults vs ~22% repaid",
       x = "Loan Status", y = "Count", fill = "Status") +
  theme(plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 12),
        legend.position = "bottom")
```

> **Important:** The 78/22 class imbalance means **accuracy alone is misleading**. We will rely on **AUC-ROC** for classification evaluation.

## Bivariate Analysis Credit Risk Exploration

### Income vs Credit Score by Loan Status

```{r scatter-income-credit, fig.width=12, fig.height=7}
ggplot(df %>% filter(person_income < 200000),
       aes(x = credit_score, y = person_income, color = factor(loan_status))) +
  geom_point(alpha = 0.3, size = 1.2) +
  scale_color_manual(values = c("0" = "#e74c3c", "1" = "#2980b9"),
                     labels = c("Default", "Repaid")) +
  scale_y_continuous(labels = scales::comma) +
  facet_wrap(~loan_status,
             labeller = labeller(loan_status = c("0" = "Default", "1" = "Repaid"))) +
  labs(title = "Income vs Credit Score by Loan Status",
       subtitle = "Higher credit scores and incomes tend toward repayment",
       x = "Credit Score", y = "Annual Income (USD)") +
  theme(plot.title = element_text(size = 16, face = "bold"),
        strip.text = element_text(size = 13, face = "bold"),
        legend.position = "none")
```

### Default Rate by Categorical Variables

```{r default-rates, fig.height=10, fig.width=12}
default_plots <- lapply(cat_vars, function(var) {
  df %>%
    group_by(.data[[var]]) %>%
    summarise(default_rate = mean(loan_status == 0), .groups = "drop") %>%
    ggplot(aes(x = reorder(.data[[var]], default_rate), y = default_rate)) +
    geom_col(fill = "steelblue", alpha = 0.8) +
    geom_text(aes(label = paste0(round(default_rate * 100, 1), "%")),
              hjust = -0.1, size = 3) +
    coord_flip() +
    labs(title = paste("Default Rate by", var), x = var, y = "Default Rate") +
    scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
    theme(plot.title = element_text(size = 10, face = "bold"))
})

do.call(grid.arrange, c(default_plots, ncol = 2))
```

### Loan Amount vs Income by Default Status

```{r loan-income-scatter, fig.width=12, fig.height=7}
ggplot(df %>% filter(person_income < 200000),
       aes(x = person_income, y = loan_amnt, color = factor(loan_status))) +
  geom_point(alpha = 0.3, size = 1.2) +
  scale_color_manual(values = c("0" = "#e74c3c", "1" = "#2980b9"),
                     labels = c("Default", "Repaid")) +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  facet_wrap(~loan_status,
             labeller = labeller(loan_status = c("0" = "Default", "1" = "Repaid"))) +
  labs(title = "Loan Amount vs Income by Loan Status",
       subtitle = "Defaults concentrate at higher loan amounts relative to income",
       x = "Annual Income (USD)", y = "Loan Amount (USD)") +
  theme(plot.title = element_text(size = 16, face = "bold"),
        strip.text = element_text(size = 13, face = "bold"),
        legend.position = "none")
```

## Correlation Analysis

```{r correlation-matrix, fig.width=10, fig.height=8}
numeric_df <- df %>% dplyr::select(where(is.numeric))
cor_matrix  <- cor(numeric_df, use = "complete.obs")

ggcorrplot(cor_matrix, method = "square", type = "lower",
           lab = TRUE, lab_size = 3,
           colors = c("#e74c3c", "white", "#2980b9"),
           title = "Correlation Matrix of Numeric Variables",
           ggtheme = theme_classic())
```

```{r corrplot-alt, fig.width=8, fig.height=8}
corrplot(cor_matrix, method = "color", type = "upper",
         addCoef.col = "black", number.cex = 0.7,
         tl.col = "black", tl.srt = 45,
         title = "Correlation Plot", mar = c(0, 0, 2, 0))
```

**Correlation Insights:**

-   `loan_percent_income` has the strongest positive correlation with `loan_status` the most influential default predictor.
-   `credit_score` is negatively correlated with default (higher score - less default).
-   `person_income` and `loan_amnt` show moderate positive correlation (higher earners request larger loans).

------------------------------------------------------------------------

# Data Preprocessing

## Outlier Treatment

```{r outlier-treatment}
cat("Before outlier treatment:", nrow(df), "rows\n\n")

cat("Ages > 80:", sum(df$person_age > 80), "\n")
df <- df %>% filter(person_age <= 80)

cat("Employment experience > 60:", sum(df$person_emp_exp > 60), "\n")
df <- df %>% filter(person_emp_exp <= 60)

income_cap <- quantile(df$person_income, 0.995)
cat("Income cap (99.5th percentile):", income_cap, "\n")
cat("Records above cap:", sum(df$person_income > income_cap), "\n")
df <- df %>% filter(person_income <= income_cap)

cat("\nAfter outlier treatment:", nrow(df), "rows\n")
```

## Encode Categorical Variables

```{r encode-categoricals}
df <- df %>%
  mutate(
    person_gender                  = as.factor(person_gender),
    person_education               = factor(person_education,
                                            levels = c("High School", "Associate",
                                                       "Bachelor", "Master", "Doctorate"),
                                            ordered = TRUE),
    person_home_ownership          = as.factor(person_home_ownership),
    loan_intent                    = as.factor(loan_intent),
    previous_loan_defaults_on_file = as.factor(previous_loan_defaults_on_file),
    loan_status                    = as.factor(loan_status)
  )

str(df)
```

------------------------------------------------------------------------

# Feature Engineering

We create **domain-informed features** that capture relationships the raw variables miss.

| New Feature | Formula / Logic | Rationale |
|-----------------------|------------------------------|--------------------|
| `loan_credit_ratio` | loan_amnt / credit_score | Loan burden relative to creditworthiness |
| `age_exp_ratio` | emp_exp / age | Career progression proxy |
| `income_per_exp` | income / emp_exp | Earning efficiency |
| `credit_hist_per_age` | cred_hist_length / age | Financial maturity |
| `high_risk` | loan% \> 0.3 AND credit \< 600 | Binary risk flag |
| `age_group` | Binned age | Life-stage segmentation |
| `income_bracket` | Quartile-based bins | Socioeconomic stratum |

```{r feature-engineering}
df$loan_credit_ratio   <- df$loan_amnt / df$credit_score
df$age_exp_ratio       <- ifelse(df$person_age > 0, df$person_emp_exp / df$person_age, 0)
df$income_per_exp      <- ifelse(df$person_emp_exp > 0,
                                 df$person_income / df$person_emp_exp,
                                 df$person_income)
df$credit_hist_per_age <- df$cb_person_cred_hist_length / df$person_age
df$high_risk           <- as.factor(
  ifelse(df$loan_percent_income > 0.3 & df$credit_score < 600, 1, 0))
df$age_group           <- cut(df$person_age,
                              breaks = c(19, 25, 35, 50, 80),
                              labels = c("Young Adult", "Adult", "Middle Age", "Senior"))
df$income_bracket      <- cut(df$person_income,
                              breaks = quantile(df$person_income, probs = c(0, .25, .5, .75, 1)),
                              labels = c("Low", "Lower-Mid", "Upper-Mid", "High"),
                              include.lowest = TRUE)
```

```{r feature-viz, fig.height=8, fig.width=12}
p1 <- ggplot(df, aes(x = loan_status, y = loan_credit_ratio, fill = loan_status)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("#e74c3c", "#27ae60")) +
  labs(title = "Loan-to-Credit Ratio by Status", y = "Loan / Credit Ratio") +
  theme(legend.position = "none")

p2 <- ggplot(df, aes(x = high_risk, fill = loan_status)) +
  geom_bar(position = "fill", alpha = 0.8) +
  scale_fill_manual(values = c("#e74c3c", "#27ae60"), labels = c("Default", "Repaid")) +
  labs(title = "Default Rate: High Risk vs Not", y = "Proportion") +
  theme(legend.position = "bottom")

p3 <- ggplot(df, aes(x = age_group, fill = loan_status)) +
  geom_bar(position = "fill", alpha = 0.8) +
  scale_fill_manual(values = c("#e74c3c", "#27ae60"), labels = c("Default", "Repaid")) +
  labs(title = "Default Rate by Age Group", y = "Proportion") +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 30, hjust = 1))

p4 <- ggplot(df, aes(x = income_bracket, fill = loan_status)) +
  geom_bar(position = "fill", alpha = 0.8) +
  scale_fill_manual(values = c("#e74c3c", "#27ae60"), labels = c("Default", "Repaid")) +
  labs(title = "Default Rate by Income Bracket", y = "Proportion") +
  theme(legend.position = "bottom")

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

**Key insights from engineered features:**

-   Defaulters exhibit a **higher loan-to-credit ratio** : they borrow more relative to their creditworthiness.
-   The **high-risk flag** sharply separates default rates: flagged applicants default far more often.
-   **Young adults** show the highest default proportion; **seniors** the lowest.
-   **Low-income brackets** default more frequently than high-income brackets.

------------------------------------------------------------------------

# Part A.Regression: Predicting Income {.tabset}

## Objective

Train three regression models on `person_income` using demographic and engineered features. **Linear regression is the primary model**; its OLS assumptions are checked and transformations are applied as needed.

## Prepare Regression Data

```{r regression-data-prep}
reg_df <- df %>%
  select(person_income, person_age, person_gender, person_education,
         person_emp_exp, person_home_ownership,
         age_exp_ratio, credit_hist_per_age) %>%
  drop_na()

reg_df$person_education <- factor(reg_df$person_education, ordered = FALSE)

cat("Regression dataset:", nrow(reg_df), "rows x", ncol(reg_df), "columns\n")
summary(reg_df$person_income)
```

## Income Distribution & Transformation Decision

```{r income-distribution, fig.width=14, fig.height=5}
p1 <- ggplot(reg_df, aes(x = person_income)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white", alpha = 0.8) +
  labs(title = "Raw Income",
       subtitle = paste("Skewness:", round(skewness(reg_df$person_income), 3)),
       x = "Income", y = "Count")

p2 <- ggplot(reg_df, aes(x = log(person_income))) +
  geom_histogram(bins = 50, fill = "#27ae60", color = "white", alpha = 0.8) +
  labs(title = "Log-Transformed Income",
       subtitle = paste("Skewness:", round(skewness(log(reg_df$person_income)), 3)),
       x = "log(Income)", y = "Count")

p3 <- ggplot(reg_df, aes(x = sqrt(person_income))) +
  geom_histogram(bins = 50, fill = "#e67e22", color = "white", alpha = 0.8) +
  labs(title = "Sqrt-Transformed Income",
       subtitle = paste("Skewness:", round(skewness(sqrt(reg_df$person_income)), 3)),
       x = "sqrt(Income)", y = "Count")

grid.arrange(p1, p2, p3, ncol = 3)
```

> **Decision:** The **log transformation** substantially reduces skewness, bringing the distribution closer to normal. We proceed with `log(person_income)` as the response.

```{r transform-income}
reg_df <- reg_df %>% mutate(log_income = log(person_income))
```

## TrainTest Split

```{r train-test-split-reg}
set.seed(42)
train_index <- createDataPartition(reg_df$log_income, p = 0.8, list = FALSE)
train_reg   <- reg_df[train_index, ]
test_reg    <- reg_df[-train_index, ]

cat("Training set:", nrow(train_reg), "rows | Test set:", nrow(test_reg), "rows\n")
```

## Model 1. Linear Regression (OLS)

### Full Model

```{r linear-regression-full}
lm_full <- lm(log_income ~ person_age + person_gender + person_education +
                 person_emp_exp + person_home_ownership +
                 age_exp_ratio + credit_hist_per_age,
               data = train_reg)

summary(lm_full)
```

### Stepwise Variable Selection (AIC)

```{r stepwise-selection}
lm_step <- stepAIC(lm_full, direction = "both", trace = FALSE)
summary(lm_step)
```

```{r stepwise-comparison}
cat("Full Model AIC:", AIC(lm_full), "\n")
cat("Stepwise Model AIC:", AIC(lm_step), "\n")
cat("Full Model Adj R²:", round(summary(lm_full)$adj.r.squared, 4), "\n")
cat("Stepwise Model Adj R²:", round(summary(lm_step)$adj.r.squared, 4), "\n")
```

### Checking OLS Assumptions

#### Assumption 1.Linearity (Residuals vs Fitted)

```{r assumption-linearity, fig.width=8, fig.height=5}
plot(lm_step, which = 1, main = "Residuals vs Fitted — Linearity Check")
```

> A roughly **horizontal red line centered at zero** confirms linearity. Systematic curvature would indicate violation.

#### Assumption 2.Normality of Residuals

```{r assumption-normality, fig.width=12, fig.height=5}
par(mfrow = c(1, 2))
plot(lm_step, which = 2, main = "Q-Q Plot — Normality Check")
hist(residuals(lm_step), breaks = 50, col = "steelblue", border = "white",
     main = "Histogram of Residuals", xlab = "Residuals", probability = TRUE)
curve(dnorm(x, mean = mean(residuals(lm_step)), sd = sd(residuals(lm_step))),
      add = TRUE, col = "red", lwd = 2)
par(mfrow = c(1, 1))
```

> Points closely following the diagonal in the Q-Q plot confirm approximate normality. The histogram overlay reinforces this.

#### Assumption 3.Homoscedasticity (Constant Variance)

```{r assumption-homoscedasticity, fig.width=8, fig.height=5}
plot(lm_step, which = 3, main = "Scale-Location — Homoscedasticity Check")
```

> A **roughly horizontal line with equally spread points** supports constant variance. A funnel shape would indicate heteroscedasticity.

#### Assumption 4.No Multicollinearity (VIF)

```{r vif-check}
vif_values <- vif(lm_step)
print(vif_values)

cat("\nVariables with GVIF^(1/(2*Df)) > 2.5 (potential multicollinearity):\n")
gvif_adj <- vif_values[, "GVIF^(1/(2*Df))"]
high_vif <- gvif_adj[gvif_adj > 2.5]
if (length(high_vif) > 0) print(high_vif) else cat("None detected.\n")
```

> **Rule of thumb:** GVIF\^(1/(2×Df)) \> 2.5 suggests problematic multicollinearity.

#### Influence and Leverage (Cook's Distance)

```{r cooks-distance, fig.width=8, fig.height=5}
plot(lm_step, which = 4, main = "Cook's Distance — Influential Observations")

cooks_d      <- cooks.distance(lm_step)
influential  <- which(cooks_d > 4 / nrow(train_reg))
cat("Influential observations (Cook's D > 4/n):", length(influential), "\n")
```

### All Diagnostic Plots

```{r all-diagnostics, fig.width=12, fig.height=10}
autoplot(lm_step, which = 1:6, ncol = 3) + theme_classic()
```

### Comparing Transformations: Log vs Sqrt

```{r compare-transformations}
lm_sqrt <- lm(sqrt(person_income) ~ person_age + person_gender + person_education +
                person_emp_exp + person_home_ownership,
              data = train_reg)

cat("Log Transform Adj R²:", round(summary(lm_step)$adj.r.squared, 4), "\n")
cat("Sqrt Transform Adj R²:", round(summary(lm_sqrt)$adj.r.squared, 4), "\n")
```

```{r side-by-side-diagnostics, fig.height=10, fig.width=14}
par(mar = c(5, 5, 4, 2), mfrow = c(2, 2))
plot(lm_step, which = 1, main = "Log: Residuals vs Fitted", sub.caption = "", caption = "")
plot(lm_step, which = 2, main = "Log: Q-Q Plot", sub.caption = "", caption = "")
plot(lm_sqrt, which = 1, main = "Sqrt: Residuals vs Fitted", sub.caption = "", caption = "")
plot(lm_sqrt, which = 2, main = "Sqrt: Q-Q Plot", sub.caption = "", caption = "")
par(mfrow = c(1, 1))
```

```{r residual-histograms, fig.height=6, fig.width=14}
par(mfrow = c(1, 2))
hist(residuals(lm_step), breaks = 50, col = "steelblue", border = "white",
     main = paste("Log Residuals — Skewness:", round(skewness(residuals(lm_step)), 3)),
     xlab = "Residuals", probability = TRUE)
curve(dnorm(x, mean(residuals(lm_step)), sd(residuals(lm_step))),
      add = TRUE, col = "red", lwd = 2)

hist(residuals(lm_sqrt), breaks = 50, col = "#e67e22", border = "white",
     main = paste("Sqrt Residuals — Skewness:", round(skewness(residuals(lm_sqrt)), 3)),
     xlab = "Residuals", probability = TRUE)
curve(dnorm(x, mean(residuals(lm_sqrt)), sd(residuals(lm_sqrt))),
      add = TRUE, col = "red", lwd = 2)
par(mfrow = c(1, 1))
```

> **Conclusion:** The **log transformation** yields better normality, lower skewness in residuals and a higher Adjusted R² it remains the preferred transformation.

## Model 2.Decision Tree Regressor

```{r decision-tree-reg}
dt_model <- rpart(person_income ~ person_age + person_gender + person_education +
                    person_emp_exp + person_home_ownership +
                    age_exp_ratio + credit_hist_per_age,
                  data = train_reg, method = "anova",
                  control = rpart.control(cp = 0.001, maxdepth = 10, minsplit = 20))

printcp(dt_model)
```

## Model 3.Random Forest Regressor

```{r random-forest-reg}
set.seed(42)
rf_model <- randomForest(person_income ~ person_age + person_gender +
                           person_education + person_emp_exp +
                           person_home_ownership + age_exp_ratio +
                           credit_hist_per_age,
                         data = train_reg, ntree = 500, mtry = 3, importance = TRUE)

print(rf_model)
```

```{r rf-importance, fig.width=10, fig.height=6}
varImpPlot(rf_model, main = "Random Forest — Variable Importance (Regression)")
```

```{r rf-convergence, fig.width=8, fig.height=5}
plot(rf_model, main = "Random Forest — Error Convergence")
```

## Regression Model Comparison

```{r regression-predictions}
pred_lm <- exp(predict(lm_step, newdata = test_reg))
pred_dt <- predict(dt_model, newdata = test_reg)
pred_rf <- predict(rf_model, newdata = test_reg)
actual  <- test_reg$person_income
```

```{r regression-metrics}
calc_metrics <- function(actual, predicted, model_name) {
  data.frame(
    Model     = model_name,
    RMSE      = round(rmse(actual, predicted), 2),
    MAE       = round(mae(actual, predicted), 2),
    R_Squared = round(1 - sum((actual - predicted)^2) / sum((actual - mean(actual))^2), 4),
    MAPE      = round(mean(abs((actual - predicted) / actual)) * 100, 2)
  )
}

comparison_reg <- rbind(
  calc_metrics(actual, pred_lm, "Linear Regression (log)"),
  calc_metrics(actual, pred_dt, "Decision Tree"),
  calc_metrics(actual, pred_rf, "Random Forest")
)

print(comparison_reg)
```

```{r regression-comparison-viz, fig.width=12, fig.height=5}
comparison_long <- comparison_reg %>%
  pivot_longer(cols = c(RMSE, MAE, R_Squared), names_to = "Metric", values_to = "Value")

ggplot(comparison_long, aes(x = Model, y = Value, fill = Model)) +
  geom_col(alpha = 0.8) +
  facet_wrap(~Metric, scales = "free_y") +
  labs(title = "Regression Model Comparison", y = "Value") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1), legend.position = "none")
```

### Actual vs Predicted

```{r actual-vs-predicted, fig.width=16, fig.height=6}
pred_df <- data.frame(
  Actual    = rep(actual, 3),
  Predicted = c(pred_lm, pred_dt, pred_rf),
  Model     = rep(c("Linear Regression", "Decision Tree", "Random Forest"),
                  each = length(actual))
)

ggplot(pred_df, aes(x = Actual, y = Predicted)) +
  geom_point(alpha = 0.15, size = 1, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed", linewidth = 1) +
  facet_wrap(~Model, ncol = 3) +
  scale_x_continuous(labels = scales::comma, limits = c(0, 300000)) +
  scale_y_continuous(labels = scales::comma, limits = c(0, 300000)) +
  labs(title = "Actual vs Predicted Income",
       subtitle = "Points on the red line = perfect prediction",
       x = "Actual Income (USD)", y = "Predicted Income (USD)") +
  theme(plot.title = element_text(size = 16, face = "bold"),
        strip.text = element_text(size = 13, face = "bold"))
```

### Best Regression Model

```{r regression-conclusion}
best_model_idx <- which.max(comparison_reg$R_Squared)
cat("═══════════════════════════════════════\n")
cat("  BEST REGRESSION MODEL\n")
cat("═══════════════════════════════════════\n")
cat("  Model    :", comparison_reg$Model[best_model_idx], "\n")
cat("  R-squared:", comparison_reg$R_Squared[best_model_idx], "\n")
cat("  RMSE     :", comparison_reg$RMSE[best_model_idx], "\n")
cat("  MAE      :", comparison_reg$MAE[best_model_idx], "\n")
cat("═══════════════════════════════════════\n")
```

------------------------------------------------------------------------

# Part B.Classification: Loan Default Prediction {.tabset}

## Objective

Predict whether a loan applicant will **default** (`loan_status = 0`) or **repay** (`loan_status = 1`) using all available features, including engineered ones.

## Prepare Classification Data

```{r classification-data-prep}
clf_df <- df %>%
  dplyr::select(loan_status, person_age, person_gender, person_education, person_income,
                person_emp_exp, person_home_ownership, loan_amnt, loan_intent,
                loan_int_rate, loan_percent_income, cb_person_cred_hist_length,
                credit_score, previous_loan_defaults_on_file,
                loan_credit_ratio, age_exp_ratio, high_risk) %>%
  drop_na()

clf_df$person_education <- factor(clf_df$person_education, ordered = FALSE)

cat("Classification dataset:", nrow(clf_df), "rows x", ncol(clf_df), "columns\n")
```

## TrainTest Split

```{r train-test-split-clf}
set.seed(42)
train_idx_clf <- createDataPartition(clf_df$loan_status, p = 0.8, list = FALSE)
train_clf     <- clf_df[train_idx_clf, ]
test_clf      <- clf_df[-train_idx_clf, ]

cat("Training:", nrow(train_clf), "| Test:", nrow(test_clf), "\n")
```

## Model 1.Logistic Regression

Logistic regression is the natural extension of linear regression for binary outcomes and serves as our baseline classifier.

```{r logistic-regression}
log_model <- glm(loan_status ~ ., data = train_clf, family = binomial(link = "logit"))
log_step  <- stepAIC(log_model, direction = "both", trace = FALSE)

prob_log  <- predict(log_step, newdata = test_clf, type = "response")
pred_log  <- factor(ifelse(prob_log > 0.5, 1, 0), levels = c("0", "1"))

cat("\n=== Logistic Regression ===\n")
cm_log <- confusionMatrix(pred_log, test_clf$loan_status, positive = "1")
print(cm_log)
```

## Model 2.Random Forest Classifier

```{r random-forest-clf}
set.seed(42)
rf_clf <- randomForest(loan_status ~ ., data = train_clf,
                       ntree = 500, mtry = 4, importance = TRUE)

pred_rf_clf <- predict(rf_clf, newdata = test_clf)
prob_rf_clf <- predict(rf_clf, newdata = test_clf, type = "prob")[, 2]

cat("\n=== Random Forest ===\n")
cm_rf <- confusionMatrix(pred_rf_clf, test_clf$loan_status, positive = "1")
print(cm_rf)
```

## Model 3.Decision Tree Classifier

```{r decision-tree-clf}
dt_clf <- rpart(loan_status ~ ., data = train_clf, method = "class",
                control = rpart.control(cp = 0.001, maxdepth = 8))

pred_dt_clf <- predict(dt_clf, newdata = test_clf, type = "class")
prob_dt_clf <- predict(dt_clf, newdata = test_clf, type = "prob")[, 2]

cat("\n=== Decision Tree ===\n")
cm_dt <- confusionMatrix(pred_dt_clf, test_clf$loan_status, positive = "1")
print(cm_dt)
```

```{r dt-tree-viz, fig.width=14, fig.height=8}
rpart.plot(dt_clf, type = 4, extra = 104, roundint = FALSE,
           main = "Decision Tree — Loan Default Classification",
           box.palette = "RdGn", shadow.col = "gray")
```

## ROC Curves & AUC Comparison

**Why AUC over Accuracy?** With 78 % defaults, a naive "always predict default" model achieves 78 % accuracy but is useless. AUC measures how well the model **separates** the two classes regardless of proportion.

**ROC** stands for **Receiver Operating Characteristic**. It's a curve that plots two things as you change the classification threshold (the 0.5 cutoff).\
**AUC** stands for **Area Under the Curve**. It's a single number that summarizes the entire ROC curve:

-   **AUC = 1.0** perfect model, never makes mistakes

-   **AUC = 0.9–1.0** excellent model

-   **AUC = 0.8–0.9** good model

```{r roc-curves, fig.width=9, fig.height=7}
roc_log <- roc(test_clf$loan_status, prob_log, levels = c("0", "1"), direction = "<")
roc_rf  <- roc(test_clf$loan_status, prob_rf_clf, levels = c("0", "1"), direction = "<")
roc_dt  <- roc(test_clf$loan_status, prob_dt_clf, levels = c("0", "1"), direction = "<")

plot(roc_log, col = "#2980b9", lwd = 2, main = "ROC Curves — Classification Models")
plot(roc_rf, col = "#27ae60", lwd = 2, add = TRUE)
plot(roc_dt, col = "#e74c3c", lwd = 2, add = TRUE)
abline(a = 0, b = 1, lty = 2, col = "gray50")

legend("bottomright",
       legend = c(
         paste("Logistic Regression (AUC =", round(auc(roc_log), 4), ")"),
         paste("Random Forest       (AUC =", round(auc(roc_rf), 4), ")"),
         paste("Decision Tree       (AUC =", round(auc(roc_dt), 4), ")")
       ),
       col = c("#2980b9", "#27ae60", "#e74c3c"), lwd = 2, cex = 0.9)
```

## Classification Summary Table

```{r classification-summary}
clf_comparison <- data.frame(
  Model       = c("Logistic Regression", "Random Forest", "Decision Tree"),
  Accuracy    = c(cm_log$overall["Accuracy"], cm_rf$overall["Accuracy"],
                  cm_dt$overall["Accuracy"]),
  Sensitivity = c(cm_log$byClass["Sensitivity"], cm_rf$byClass["Sensitivity"],
                  cm_dt$byClass["Sensitivity"]),
  Specificity = c(cm_log$byClass["Specificity"], cm_rf$byClass["Specificity"],
                  cm_dt$byClass["Specificity"]),
  F1_Score    = c(cm_log$byClass["F1"], cm_rf$byClass["F1"], cm_dt$byClass["F1"]),
  AUC         = c(auc(roc_log), auc(roc_rf), auc(roc_dt))
) %>%
  mutate(across(where(is.numeric), ~ round(., 4)))

print(clf_comparison)
```

```{r classification-viz, fig.width=14, fig.height=5}
clf_long <- clf_comparison %>%
  pivot_longer(cols = -Model, names_to = "Metric", values_to = "Value")

ggplot(clf_long, aes(x = Model, y = Value, fill = Model)) +
  geom_col(alpha = 0.8) +
  facet_wrap(~Metric, scales = "free_y", ncol = 5) +
  labs(title = "Classification Model Performance Comparison", y = "Score") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1), legend.position = "none") +
  scale_fill_manual(values = c("#2980b9", "#e74c3c", "#27ae60"))
```

### Best Classification Model

```{r classification-conclusion}
best_clf_idx <- which.max(clf_comparison$AUC)
cat("═══════════════════════════════════════\n")
cat("  BEST CLASSIFICATION MODEL\n")
cat("═══════════════════════════════════════\n")
cat("  Model   :", clf_comparison$Model[best_clf_idx], "\n")
cat("  AUC     :", clf_comparison$AUC[best_clf_idx], "\n")
cat("  Accuracy:", clf_comparison$Accuracy[best_clf_idx], "\n")
cat("  F1 Score:", clf_comparison$F1_Score[best_clf_idx], "\n")
cat("═══════════════════════════════════════\n")
```

------------------------------------------------------------------------

# Credit Risk Analysis

This section synthesizes findings on how **income**, **credit score**, and **prior defaults** interact to drive loan outcomes.

```{r credit-risk-summary}
cat("══════════════════════════════════════════════\n")
cat("           CREDIT RISK INSIGHTS\n")
cat("══════════════════════════════════════════════\n")
cat("1. Higher credit scores → lower default rates.\n")
cat("2. loan_percent_income is the strongest predictor of default.\n")
cat("3. Previous loan defaults strongly predict future defaults.\n")
cat("4. High-risk applicants (high loan %, low credit) default\n")
cat("   at substantially higher rates.\n")
cat("5. Income alone is a weak predictor; its ratio to loan\n")
cat("   amount matters far more.\n")
cat("══════════════════════════════════════════════\n")
```

```{r credit-risk-viz, fig.width=14, fig.height=6}
p_cr1 <- ggplot(df, aes(x = credit_score, fill = loan_status)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("#e74c3c", "#27ae60"), labels = c("Default", "Repaid")) +
  labs(title = "Credit Score Distribution by Loan Status",
       x = "Credit Score", y = "Density", fill = "Status") +
  theme(legend.position = "bottom")

p_cr2 <- ggplot(df, aes(x = previous_loan_defaults_on_file, fill = loan_status)) +
  geom_bar(position = "fill", alpha = 0.8) +
  scale_fill_manual(values = c("#e74c3c", "#27ae60"), labels = c("Default", "Repaid")) +
  labs(title = "Default Rate by Previous Defaults on File",
       x = "Previous Defaults", y = "Proportion", fill = "Status") +
  theme(legend.position = "bottom")

grid.arrange(p_cr1, p_cr2, ncol = 2)
```

------------------------------------------------------------------------

# Summary of Findings

## Regression (Income Prediction)

```{r final-regression-summary}
cat("Best Regression Model :", comparison_reg$Model[best_model_idx], "\n")
cat("Test R²              :", comparison_reg$R_Squared[best_model_idx], "\n")
cat("Test RMSE            :", comparison_reg$RMSE[best_model_idx], "\n\n")

cat("Key regression findings:\n")
cat(" - Log transformation of income was necessary to meet normality assumptions.\n")
cat(" - Stepwise AIC selection (direction = 'both') identified optimal predictors.\n")
cat(" - All four OLS assumptions were systematically checked:\n")
cat("     Linearity, Normality, Homoscedasticity, Multicollinearity.\n")
cat(" - Log outperformed sqrt transformation based on residual diagnostics.\n")
```

## Classification (Loan Default Prediction)

```{r final-classification-summary}
cat("Best Classification Model :", clf_comparison$Model[best_clf_idx], "\n")
cat("AUC                       :", clf_comparison$AUC[best_clf_idx], "\n")
cat("Accuracy                  :", clf_comparison$Accuracy[best_clf_idx], "\n\n")

cat("Key classification findings:\n")
cat(" - Target variable is imbalanced (78% default, 22% repaid).\n")
cat(" - ROC-AUC is the primary metric due to class imbalance.\n")
cat(" - Feature engineering (loan_credit_ratio, high_risk flag) added predictive value.\n")
cat(" - Logistic Regression serves as interpretable baseline;\n")
cat("   tree-based models may capture non-linear patterns.\n")
```

## Credit Risk Analysis

```{r credit-risk-final}
cat("Credit Risk Insights:\n")
cat(" - Higher credit scores are associated with lower default rates.\n")
cat(" - loan_percent_income is the single strongest predictor of default.\n")
cat(" - Previous loan defaults strongly predict future default behavior.\n")
cat(" - High-risk applicants (high loan %, low credit) default at much higher rates.\n")
```

------------------------------------------------------------------------

# Limitations & Future Work

**Limitations:**

-   The dataset has no missing values, which is unrealistic for real-world loan data.
-   Income prediction using only demographics has inherent limits occupation and industry would improve models.
-   The 78/22 class imbalance was not addressed with resampling techniques (SMOTE, oversampling).
-   Cross-validation was not performed; it would strengthen reliability of all results.

**Future Work:**

-   Implement k-fold cross-validation for more robust model evaluation.
-   Explore regularized regression (Ridge, Lasso) for income prediction.
-   Add ensemble methods (XGBoost, stacking) for classification.
-   Incorporate additional features such as occupation, industry, and geographic data.

------------------------------------------------------------------------
