---
title: "Population_Practice"
author: "Joshua_Ngumi"
date: "2026-04-14"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    fig_width: 9
    fig_height: 5
  pdf_document:
    toc: true
    toc_depth: 3
    fig_width: 7
    fig_height: 
runtime: shiny
---

## Environment Setup

```{r load-libraries}
libs <- c("dplyr","tidyr","readr","purrr", "fst", "finalfit", "shiny", "ggplot2", "DT","lmtest", "car", "MASS", "broom","lmtest", "car", "MASS")

for (lib in libs) {
  if (!(lib %in% installed.packages())) install.packages(lib, quietly = TRUE)
  library(lib, character.only = TRUE)
}

select <- dplyr::select
mutate <- dplyr::mutate
filter <- dplyr::filter
summarise <- dplyr::summarise


theme_set(theme_classic())
```

## Load the Data
#### First convert to dot fst for faster loading

```{r load data}
csv_path <- "C:/Users/Light House/Desktop/MastersDSA/semThree-DSA/Linear Models/Data_Project/population_master_practise.csv"

fst_path <- "population_master_practise.fst"

# Convert first
if (!(file.exists(fst_path))) {
  population_df <- read.csv(csv_path)
  
  write_fst(population_df, fst_path)
}

# load
population_df <- read_fst(fst_path)

head(population_df)

```

## Initial clean up

```{r validate-data-types}
glimpse(population_df)

```
## Describe & model with **finalfit**

We'll predict **life expectancy** from a handful of demographic indicators.


```{r finalfit-table}

explanatory <- c("crude_birth_rate", "crude_death_rate",
                 "total_fertility_rate", "growth_rate",
                 "infant_mortality")
dependent   <- "life_expectancy"


# Summary table (finalfit)
population_df |>
  summary_factorlist(dependent, explanatory,
                     cont = "mean", p = TRUE, add_dependent_label = TRUE)

```



```{r finalfit-model}
# Linear regression via finalfit
ff_fit <- population_df |> finalfit(dependent, explanatory, metrics = TRUE)
ff_fit[[1]]   # coefficient table
ff_fit[[2]]   # model metrics (R², AIC, etc.)

# Keep the underlying lm for the Shiny predictor below
lm_fit <- lm(reformulate(explanatory, dependent), data = population_df)



```




## Interactive prediction UI (Shiny)

```{r shiny-app, echo=FALSE}
library(shiny)
library(ggplot2)
library(tidyr)

# Make sure the objects from earlier chunks are clean
population_df <- drop_na(population_df)

rng <- function(x) {
  r <- range(x, na.rm = TRUE)
  if (any(!is.finite(r))) stop("Column is empty or all-NA")
  round(r, 2)
}
med <- function(x) median(x, na.rm = TRUE)

shinyApp(
  ui = fluidPage(
    titlePanel("Predict Life Expectancy"),
    sidebarLayout(
      sidebarPanel(
        sliderInput("cbr", "Crude birth rate",
                    min = rng(population_df$crude_birth_rate)[1],
                    max = rng(population_df$crude_birth_rate)[2],
                    value = med(population_df$crude_birth_rate), step = 0.1),
        sliderInput("cdr", "Crude death rate",
                    min = rng(population_df$crude_death_rate)[1],
                    max = rng(population_df$crude_death_rate)[2],
                    value = med(population_df$crude_death_rate), step = 0.1),
        sliderInput("tfr", "Total fertility rate",
                    min = rng(population_df$total_fertility_rate)[1],
                    max = rng(population_df$total_fertility_rate)[2],
                    value = med(population_df$total_fertility_rate), step = 0.05),
        sliderInput("gr", "Growth rate",
                    min = rng(population_df$growth_rate)[1],
                    max = rng(population_df$growth_rate)[2],
                    value = med(population_df$growth_rate), step = 0.05),
        sliderInput("imr", "Infant mortality",
                    min = rng(population_df$infant_mortality)[1],
                    max = rng(population_df$infant_mortality)[2],
                    value = med(population_df$infant_mortality), step = 0.5)
      ),
      mainPanel(
        h3("Predicted life expectancy"),
        verbatimTextOutput("pred"),
        plotOutput("plot", height = "320px")
      )
    )
  ),

  server = function(input, output, session) {
    pred <- reactive({
      newdat <- data.frame(
        crude_birth_rate     = input$cbr,
        crude_death_rate     = input$cdr,
        total_fertility_rate = input$tfr,
        growth_rate          = input$gr,
        infant_mortality     = input$imr
      )
      predict(lm_fit, newdata = newdat, interval = "prediction")
    })

    output$pred <- renderPrint({
      p <- pred()
      cat(sprintf("Point estimate : %.2f years\n", p[1, "fit"]))
      cat(sprintf("95%% PI        : [%.2f , %.2f]\n", p[1, "lwr"], p[1, "upr"]))
    })

    output$plot <- renderPlot({
      ggplot(population_df, aes(infant_mortality, life_expectancy)) +
        geom_point(alpha = 0.15) +
        geom_vline(xintercept = input$imr, colour = "steelblue", linewidth = 1) +
        geom_hline(yintercept = pred()[1, "fit"], colour = "tomato", linewidth = 1) +
        labs(x = "Infant mortality", y = "Life expectancy",
             title = "Your prediction in context") +
        theme_minimal()
    })
  },
  options = list(height = 560)
)
```

Two things that were also wrong in your last paste: `rng` was redefined (overwriting the guarded version) and the plot used `model_df` instead of `population_df`. Both are corrected above.

After saving, hit **Run Document**. The sliders should render with real numbers now.





















