---
title: "Loan Prediction"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

# Introduction

## Libraries

```{r load libraries}
libs <- c("tidyverse", "dplyr", "gridExtra", "ggplot2", "ggfortify","broom", "ggcorrplot", "corrr", "corrplot", 
          "MASS", 'datarium', 'randomForest', 'rpart', 'Hmisc', "caret", "Metrics","car","rpart.plot","e1071","pROC")

for(lib in libs){
  if(!(lib %in% installed.packages())){
    install.packages(lib)
  }
  library(lib, character.only = TRUE)
}


theme_set(theme_classic())

```

```{r}
select <- dplyr::select
```

```{r fix-conflicts}
select  <- dplyr::select
mutate  <- dplyr::mutate
filter  <- dplyr::filter
summarise <- dplyr::summarise

```

## Load Dataset

```{r load dataset}

df <-read_csv("C:/Users/Light House/Desktop/MastersDSA/semThree-DSA/Linear Models/Data_Project/loan_data.csv")

head(df)

```

# Data Understanding

## Structure and Types

```{r structure}
 

str(df)

```

```{r stats summary}

summary(df)

```

-   `person_age` has a maximum of 144, which is biologically implausible and signals outliers requiring treatment.

-   `person_income` ranges from \$8,000 to over \$7.2 million — extreme right skew.

-   `person_emp_exp` has a maximum of 125 years, another implausible outlier.

-   `loan_status` has a mean of \~0.22, meaning only 22% of loans were repaid (class imbalance: 78% defaults).

    ## Missing Values

```{r missing values}

missing_counts <- colSums(is.na(df))
cat("Missing values per column:\n")
print(missing_counts)
cat("/n Total missing values :", sum(missing_counts),"\n")

```

**Result:** No missing values exist in this dataset, so imputation is not needed. However, we still need to handle **implausible outlier values.**

## Duplicate Check

```{r duplicates}

n_duplicates <- sum (duplicated(df))
cat("Number of duplicate:", n_duplicates)

if (n_duplicates > 0){
  df <- df%>% distinct ()
  cat("After removing duplicates:", nrow(df),"rows remain.")
}

```

# Exploratory Data Analysis

## Univariate Analysis:Numeric Varibles

Here we examined the **distribution of each numeric variable using histograms and boxplots** to **identify skewness, outliers and the general shape of each distribution.**

```{r univariate-numeric, fig.height=20, fig.width=16}
# Histograms with count annotations
hist_plots <- lapply(numeric_vars, function(var) {
  ggplot(df, aes(x = .data[[var]])) +
    geom_histogram(bins = 40, fill = "steelblue", color = "white", alpha = 0.8) +
    geom_text(stat = "bin", bins = 40,
              aes(label = ifelse(after_stat(count) > 0, after_stat(count), "")),
              vjust = -0.3, size = 2.5) +
    labs(title = var, x = NULL, y = "Count") +
    scale_x_continuous(labels = scales::comma) +
    scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
    theme(
      plot.title  = element_text(size = 14, face = "bold"),
      axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
      axis.text.y = element_text(size = 10),
      axis.title.y = element_text(size = 12),
      plot.margin = ggplot2::margin(15, 20, 15, 20, unit = "pt")
    )
})

do.call(grid.arrange, c(hist_plots, ncol = 2))
```

```{r boxplots-numeric, fig.height=14, fig.width=12}
# Boxplots for outlier detection
box_plots <- lapply(numeric_vars, function(var) {
  ggplot(df, aes(y = .data[[var]])) +
    geom_boxplot(fill = "steelblue", alpha = 0.6, outlier.color = "red", 
                 outlier.size = 1) +
    labs(title = var, y = NULL) +
    scale_y_continuous(labels = scales::comma) +
    theme(
      plot.title  = element_text(size = 11, face = "bold"),
      axis.text.y = element_text(size = 7),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      plot.margin = ggplot2::margin(5, 10, 5, 10, unit = "pt")
    )
})

do.call(grid.arrange, c(box_plots, ncol = 4))

```

**Observations:**

-   `person_age`: Right-skewed with extreme outliers above \~80. Ages \> 100 are biologically impossible.
-   `person_income`: Heavily right-skewed with a very long tail-a log transformation will be essential for regression.
-   `person_emp_exp`: Contains implausible values (e.g., 125 years of experience).
-   `credit_score`: Roughly normally distributed, centered around 630.

### Skewness Assessment

```{r}

skew_table <- data.frame(
  Variable = numeric_vars,
  Skewness = sapply(numeric_vars, function(v) round(skewness(df[[v]], na.rm = TRUE), 3)),
  Kurtosis = sapply(numeric_vars, function(v) round(kurtosis(df[[v]], na.rm = TRUE), 3))
)

print(skew_table)
```

**Rule of thumb:** Skewness values beyond ±1 suggest the need for transformation in linear regression.

## Univariate Analysis: Categorical Variables

```{r univariate-categorical, fig.height=10, fig.width=12}
cat_vars <- c("person_gender", "person_education", "person_home_ownership", 
              "loan_intent", "previous_loan_defaults_on_file")

cat_plots <- lapply(cat_vars, function(var) {
  ggplot(df, aes(x = reorder(.data[[var]], .data[[var]], function(x) -length(x)))) +
    geom_bar(fill = "steelblue", alpha = 0.8) +
    geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5, size = 3) +
    labs(title = var, x = var, y = "Count") +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          plot.title = element_text(size = 10, face = "bold"))
})

do.call(grid.arrange, c(cat_plots, ncol = 3))
```

## Target Variable Distribution

```{r target-distribution, fig.width=8, fig.height=6}
ggplot(df, aes(x = factor(loan_status), fill = factor(loan_status))) +
  geom_bar(alpha = 0.8, width = 0.5) +
  geom_text(stat = "count", aes(label = paste0(after_stat(count), 
            " (", round(after_stat(count)/nrow(df)*100, 1), "%)")), 
            vjust = -0.5, size = 5, fontface = "bold") +
  scale_fill_manual(values = c("0" = "#e74c3c", "1" = "#27ae60"), 
                    labels = c("Default (0)", "Repaid (1)")) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.12))) +
  scale_x_discrete(labels = c("0" = "Default (0)", "1" = "Repaid (1)")) +
  labs(title = "Loan Status Distribution (Target Variable)",
       subtitle = "Significant class imbalance: 78% defaults vs 22% repaid",
       x = "Loan Status", y = "Count", fill = "Status") +
  theme(
    plot.title    = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.text.x   = element_text(size = 12, face = "bold"),
    axis.text.y   = element_text(size = 11),
    axis.title     = element_text(size = 13),
    legend.position = "bottom",
    legend.text     = element_text(size = 11)
  )

```

**Important:** The target is imbalanced (78/22 split). This will need to be considered when building classification models accuracy alone will be misleading.

## Bivariate Analysis: Credit Risk Exploration

## Train-Test Split

### Income vs Credit Score by Loan Status

```{r scatter-income-credit, fig.width=12, fig.height=7}
ggplot(df %>% filter(person_income < 200000), 
       aes(x = credit_score, y = person_income, color = factor(loan_status))) +
  geom_point(alpha = 0.3, size = 1.2) +
  scale_color_manual(values = c("0" = "#e74c3c", "1" = "#2980b9"),
                     labels = c("Default", "Repaid")) +
  scale_y_continuous(labels = scales::comma) +
  facet_wrap(~loan_status, labeller = labeller(loan_status = c("0" = "Default", "1" = "Repaid"))) +
  labs(title = "Income vs Credit Score by Loan Status",
       subtitle = "Higher credit scores and incomes tend toward repayment",
       x = "Credit Score", y = "Annual Income (USD)", color = "Loan Status") +
  theme(
    plot.title    = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.text     = element_text(size = 11),
    axis.title    = element_text(size = 13),
    strip.text    = element_text(size = 13, face = "bold"),
    legend.position = "none"
  )
```

### Default Rate by Categorical Variables

```{r default-rates, fig.height=10, fig.width=12}
default_plots <- lapply(cat_vars, function(var) {
  df %>%
    group_by(.data[[var]]) %>%
    summarise(default_rate = mean(loan_status == 0), .groups = "drop") %>%
    ggplot(aes(x = reorder(.data[[var]], default_rate), y = default_rate)) +
    geom_col(fill = "steelblue", alpha = 0.8) +
    geom_text(aes(label = paste0(round(default_rate * 100, 1), "%")), 
              hjust = -0.1, size = 3) +
    coord_flip() +
    labs(title = paste("Default Rate by", var), x = var, y = "Default Rate") +
    scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
    theme(plot.title = element_text(size = 10, face = "bold"))
})

do.call(grid.arrange, c(default_plots, ncol = 2))
```

### Loan Amount vs Income by Default Status

```{r loan-income-scatter, fig.width=12, fig.height=7}
ggplot(df %>% filter(person_income < 200000), 
       aes(x = person_income, y = loan_amnt, color = factor(loan_status))) +
  geom_point(alpha = 0.3, size = 1.2) +
  scale_color_manual(values = c("0" = "#e74c3c", "1" = "#2980b9"),
                     labels = c("Default", "Repaid")) +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  facet_wrap(~loan_status, labeller = labeller(loan_status = c("0" = "Default", "1" = "Repaid"))) +
  labs(title = "Loan Amount vs Income by Loan Status",
       subtitle = "Defaults are more concentrated at higher loan amounts relative to income",
       x = "Annual Income (USD)", y = "Loan Amount (USD)") +
  theme(
    plot.title    = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.text     = element_text(size = 11),
    axis.title    = element_text(size = 13),
    strip.text    = element_text(size = 13, face = "bold"),
    legend.position = "none"
  )
```

## Correlation Analysis

```{r correlation-matrix, fig.width=10, fig.height=8}
# Select numeric variables for correlation
numeric_df <- df %>% dplyr::select(where(is.numeric))

cor_matrix <- cor(numeric_df, use = "complete.obs")

ggcorrplot(cor_matrix, 
           method = "square",
           type = "lower",
           lab = TRUE,
           lab_size = 3,
           colors = c("#e74c3c", "white", "#2980b9"),
           title = "Correlation Matrix of Numeric Variables",
           ggtheme = theme_classic())

corrplot(cor_matrix, method = "color", type = "upper", 
         addCoef.col = "black", number.cex = 0.7,
         tl.col = "black", tl.srt = 45,
         title = "Correlation Plot", mar = c(0,0,2,0))

```

```{r corrplot-alternative, fig.width=8, fig.height=8}
corrplot(cor_matrix, method = "color", type = "upper", 
         addCoef.col = "black", number.cex = 0.7,
         tl.col = "black", tl.srt = 45,
         title = "Correlation Plot", mar = c(0,0,2,0))

```

**Correlation Insights:**

-   `loan_percent_income` has the strongest positive correlation with `loan_status` indicating it is the most influential predictor of loan outcome.

-   `credit_score` is negatively correlated with default (higher scores and less default).

-   `person_income` and `loan_amnt` show moderate positive correlation (people with higher incomes request larger loans).

    # Data Preprocessing

## Outlier Treatment

From our EDA, we identified biologically and logically implausible values that need to be removed or capped.

```{r outlier-treatment}
cat("Before outlier treatment:", nrow(df), "rows\n\n")

# Remove biologically implausible ages (> 80)
cat("Ages > 80:", sum(df$person_age > 80), "\n")
df <- df %>% filter(person_age <= 80)

# Remove implausible employment experience (> 60 years)
cat("Employment experience > 60:", sum(df$person_emp_exp > 60), "\n")
df <- df %>% filter(person_emp_exp <= 60)

# Cap extreme income values (> 99.5th percentile)
income_cap <- quantile(df$person_income, 0.995)
cat("Income cap (99.5th percentile):", income_cap, "\n")
cat("Records above cap:", sum(df$person_income > income_cap), "\n")
df <- df %>% filter(person_income <= income_cap)

cat("\nAfter outlier treatment:", nrow(df), "rows\n")

```

## Encode Categorical Variables

For modeling, we convert categorical variables to factors (R handles factor encoding internally for\
regression and tree-based models).

```{r encode-categoricals}
df <- df %>%
  mutate(
    person_gender    = as.factor(person_gender),
    person_education = factor(person_education, 
                              levels = c("High School", "Associate", "Bachelor", 
                                         "Master", "Doctorate"),
                              ordered = TRUE),
    person_home_ownership = as.factor(person_home_ownership),
    loan_intent      = as.factor(loan_intent),
    previous_loan_defaults_on_file = as.factor(previous_loan_defaults_on_file),
    loan_status      = as.factor(loan_status)
  )

str(df)


```

# Feature Engineering

In this caser feature engineering creates new informative variables that may capture relationships the raw data does not explicitly represent.

```{r engineered-features-viz, fig.height=8, fig.width=12}
# Create features first
df$loan_credit_ratio    <- df$loan_amnt / df$credit_score
df$age_exp_ratio        <- ifelse(df$person_age > 0, df$person_emp_exp / df$person_age, 0)
df$income_per_exp       <- ifelse(df$person_emp_exp > 0, df$person_income / df$person_emp_exp, df$person_income)
df$credit_hist_per_age  <- df$cb_person_cred_hist_length / df$person_age
df$high_risk            <- as.factor(ifelse(df$loan_percent_income > 0.3 & df$credit_score < 600, 1, 0))
df$age_group            <- cut(df$person_age, breaks = c(19, 25, 35, 50, 80),
                               labels = c("Young Adult", "Adult", "Middle Age", "Senior"))
df$income_bracket       <- cut(df$person_income,
                               breaks = quantile(df$person_income, probs = c(0, 0.25, 0.5, 0.75, 1)),
                               labels = c("Low", "Lower-Mid", "Upper-Mid", "High"),
                               include.lowest = TRUE)
```

```{r engineered-features-viz, fig.height=8, fig.width=12}
p1 <- ggplot(df, aes(x = loan_status, y = loan_credit_ratio, fill = loan_status)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("#e74c3c", "#27ae60")) +
  labs(title = "Loan-to-Credit Ratio by Status", y = "Loan/Credit Ratio") +
  theme(legend.position = "none")
```

```{r}
p2 <- ggplot(df, aes(x = high_risk, fill = loan_status)) +
  geom_bar(position = "fill", alpha = 0.8) +
  scale_fill_manual(values = c("#e74c3c", "#27ae60"), 
                    labels = c("Default", "Repaid")) +
  labs(title = "Default Rate: High Risk vs Not", y = "Proportion") +
  theme(legend.position = "bottom")

```

```{r}
p3 <- ggplot(df, aes(x = age_group, fill = loan_status)) +
  geom_bar(position = "fill", alpha = 0.8) +
  scale_fill_manual(values = c("#e74c3c", "#27ae60"),
                    labels = c("Default", "Repaid")) +
  labs(title = "Default Rate by Age Group", y = "Proportion") +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r}
p4 <- ggplot(df, aes(x = income_bracket, fill = loan_status)) +
  geom_bar(position = "fill", alpha = 0.8) +
  scale_fill_manual(values = c("#e74c3c", "#27ae60"),
                    labels = c("Default", "Repaid")) +
  labs(title = "Default Rate by Income Bracket", y = "Proportion") +
  theme(legend.position = "bottom")
```

```{r}
grid.arrange(p1, p2, p3, p4, ncol = 2)

```

# Part A: Regression Predicting Income

## Objective

We trainng three regression models to predict `person_income` using personal demographic variables: `person_age`, `person_gender`, `person_education`, `person_emp_exp`, `person_home_ownership`, and engineered features.

## Prepare Regression Data

```{r regression-data-prep}
# Select personal information variables for income prediction
reg_df <- df %>%
  select(person_income, person_age, person_gender, person_education, 
         person_emp_exp, person_home_ownership, 
         age_exp_ratio, credit_hist_per_age) %>%
  drop_na()

```

```{r}
# Convert ordered factor to unordered for regression compatibility
reg_df$person_education <- factor(reg_df$person_education, ordered = FALSE)

cat("Regression dataset:", nrow(reg_df), "rows x", ncol(reg_df), "columns\n")
summary(reg_df$person_income)
```

## Check Distribution of Target Income

```{r income-distribution, fig.width=12, fig.height=5}
p1 <- ggplot(reg_df, aes(x = person_income)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white", alpha = 0.8) +
  labs(title = "Distribution of Income Raw", 
       subtitle = paste("Skewness:", round(skewness(reg_df$person_income), 3)),
       x = "Income", y = "Count")

p2 <- ggplot(reg_df, aes(x = log(person_income))) +
  geom_histogram(bins = 50, fill = "#27ae60", color = "white", alpha = 0.8) +
  labs(title = "Distribution of log Income", 
       subtitle = paste("Skewness:", round(skewness(log(reg_df$person_income)), 3)),
       x = "log(Income)", y = "Count")

p3 <- ggplot(reg_df, aes(x = sqrt(person_income))) +
  geom_histogram(bins = 50, fill = "#e67e22", color = "white", alpha = 0.8) +
  labs(title = "Distribution of sqrt Income ", 
       subtitle = paste("Skewness:", round(skewness(sqrt(reg_df$person_income)), 3)),
       x = "sqrt(Income)", y = "Count")

grid.arrange(p1, p2, p3, ncol = 3)
```

**Decision:** The log transformation substantially reduces skewness, making the distribution closer to normal. We will use `log(person_income)` as the response variable for linear regression.

```{r transform-income}
reg_df <- reg_df %>% mutate(log_income = log(person_income))
```

## Train-Test Split

```{r train-test-split-reg}
set.seed(42)
train_index <- createDataPartition(reg_df$log_income, p = 0.8, list = FALSE)
train_reg <- reg_df[train_index, ]
test_reg  <- reg_df[-train_index, ]

cat("Training set:", nrow(train_reg), "rows\n")
cat("Test set:", nrow(test_reg), "rows\n")
```

## Model 1: Linear Regression

```{r linear-regression}
lm_full <- lm(log_income ~ person_age + person_gender + person_education + 
                 person_emp_exp + person_home_ownership + age_exp_ratio + 
                 credit_hist_per_age, 
               data = train_reg)

summary(lm_full)
```

### Stepwise Variable Selection

Stepwise selection uses AIC ) to iteratively add and remove predictors at each step.

```{r stepwise-selection}
lm_step <- stepAIC(lm_full, direction = "both", trace = TRUE)
summary(lm_step)
```

```{r stepwise-comparison}
cat("Full Model AIC:", AIC(lm_full), "\n")
cat("Stepwise Model AIC:", AIC(lm_step), "\n")
cat("Full Model Adj R²:", summary(lm_full)$adj.r.squared, "\n")
cat("Stepwise Model Adj R²:", summary(lm_step)$adj.r.squared, "\n")

```

### Checking Linear Regression Assumptions

We systematically check all four key assumptions of ordinary least squares (OLS) regression

#### Assumption 1: Linearity (Residuals vs Fitted)

```{r assumption-linearity, fig.width=8, fig.height=5}
plot(lm_step, which = 1)
```

**Interpretation:** If the red line is approximately horizontal and centered at zero, the linearity assumption holds. Systematic patterns (curves, funnels) indicate non-linearity.

#### Assumption 2: Normality of Residuals

```{r assumption-normality, fig.width=12, fig.height=5}
par(mfrow = c(1, 2))
# Q-Q Plot
plot(lm_step, which = 2)

# Histogram of residuals
hist(residuals(lm_step), breaks = 50, col = "steelblue", border = "white",
     main = "Histogram of Residuals", xlab = "Residuals", probability = TRUE)
curve(dnorm(x, mean = mean(residuals(lm_step)), sd = sd(residuals(lm_step))), 
      add = TRUE, col = "red", lwd = 2)
par(mfrow = c(1, 1))
```

#### Assumption 3: Homoscedasticity Constant Variance

```{r assumption-homoscedasticity, fig.width=8, fig.height=5}
plot(lm_step, which = 3)
```

**Interpretation:** The Scale-Location plot should show a roughly horizontal line with equally spread points. A funnel shape indicates heteroscedasticity.\

#### Assumption 4: No Multicollinearity

```{r vif-check}

vif_values <- vif(lm_step)
print(vif_values)

cat("\nVariables with VIF > 5 (potential multicollinearity concern):\n")
high_vif <- vif_values[apply(as.matrix(vif_values[, "GVIF^(1/(2*Df))"]), 1, 
                             function(x) x > 2.5)]
if (length(high_vif) > 0) print(high_vif) else cat("None detected.\n")
```

**Rule of thumb:** Generalized VIF\^(1/(2\*Df)) \> 2.5 suggests multicollinearity.

#### Influence & Leverage (Cook's Distance)

```{r cooks-distance, fig.width=8, fig.height=5}
plot(lm_step, which = 4)

```

```{r influence-summary}
# Identify influential observations
cooks_d <- cooks.distance(lm_step)
influential <- which(cooks_d > 4 / nrow(train_reg))
cat("Number of influential observations (Cook's D > 4/n):", length(influential), "\n")
```

### All Diagnostic Plots Together

```{r all-diagnostics, fig.width=12, fig.height=10}
autoplot(lm_step, which = 1:6, ncol = 3) + theme_classic()

```

### Attempting Additional Transformations

If assumptions are violated, we try alternative transformations.

```{r alternative-transformations}
# Square root transformation
lm_sqrt <- lm(sqrt(person_income) ~ person_age + person_gender + person_education + 
                person_emp_exp + person_home_ownership,
              data = train_reg)

cat("Log Transform Adj R²:", summary(lm_step)$adj.r.squared, "\n")
cat("Sqrt Transform Adj R²:", summary(lm_sqrt)$adj.r.squared, "\n")
```

```{r alternative-transformations, fig.height=16, fig.width=14}
# Square root transformation
lm_sqrt <- lm(sqrt(person_income) ~ person_age + person_gender + person_education + 
                person_emp_exp + person_home_ownership,
              data = train_reg)

cat("--- Log Transform Adj R²:", summary(lm_step)$adj.r.squared, "---\n")
cat("--- Sqrt Transform Adj R²:", summary(lm_sqrt)$adj.r.squared, "---\n")

# Diagnostic plots side by side
par(mfrow = c(4, 2))

# Row 1: Residuals vs Fitted
plot(lm_step, which = 1, main = "Log Transform: Residuals vs Fitted")
plot(lm_sqrt, which = 1, main = "Sqrt Transform: Residuals vs Fitted")

# Row 2: Q-Q Plot (Normality)
plot(lm_step, which = 2, main = "Log Transform: Q-Q Plot")
plot(lm_sqrt, which = 2, main = "Sqrt Transform: Q-Q Plot")

# Row 3: Scale-Location (Homoscedasticity)
plot(lm_step, which = 3, main = "Log Transform: Scale-Location")
plot(lm_sqrt, which = 3, main = "Sqrt Transform: Scale-Location")

# Row 4: Cook's Distance (Influence)
plot(lm_step, which = 4, main = "Log Transform: Cook's Distance")
plot(lm_sqrt, which = 4, main = "Sqrt Transform: Cook's Distance")

par(mfrow = c(1, 1))

```

```{r transformation-histograms, fig.height=6, fig.width=14}
# Compare residual distributions
par(mfrow = c(1, 2))

hist(residuals(lm_step), breaks = 50, col = "steelblue", border = "white",
     main = paste("Log Transform Residuals\nSkewness:", 
                  round(skewness(residuals(lm_step)), 3)),
     xlab = "Residuals", probability = TRUE)
curve(dnorm(x, mean = mean(residuals(lm_step)), sd = sd(residuals(lm_step))),
      add = TRUE, col = "red", lwd = 2)

hist(residuals(lm_sqrt), breaks = 50, col = "#e67e22", border = "white",
     main = paste("Sqrt Transform Residuals\nSkewness:", 
                  round(skewness(residuals(lm_sqrt)), 3)),
     xlab = "Residuals", probability = TRUE)
curve(dnorm(x, mean = mean(residuals(lm_sqrt)), sd = sd(residuals(lm_sqrt))),
      add = TRUE, col = "red", lwd = 2)

par(mfrow = c(1, 1))
```

-   **Row 1 (Residuals vs Fitted):** checks linearity,which transformation produces a flatter, more random scatter?

<!-- -->

-   **Row 2 (Q-Q Plot):** checks normality,which one follows the diagonal line more closely

<!-- -->

-   **Row 3 (Scale-Location):** checks homoscedasticity,which one has a more horizontal red line

<!-- -->

-   **Row 4 (Cook's Distance):** checks influence — which one has fewer extreme outliers?

<!-- -->

-   **Histogram chunk:** directly compares the residual distributions with skewness values and a normal curve overlay

## Model 2: Decision Tree Regressor

```{r decision-tree-reg}
dt_model <- rpart(person_income ~ person_age + person_gender + person_education + 
                    person_emp_exp + person_home_ownership + age_exp_ratio + 
                    credit_hist_per_age,
                  data = train_reg, 
                  method = "anova",
                  control = rpart.control(cp = 0.001, maxdepth = 10, minsplit = 20))

# Print tree summary
printcp(dt_model)
```

## Model 3: Random Forest Regressor

```{r random-forest-reg}
set.seed(42)
rf_model <- randomForest(person_income ~ person_age + person_gender + 
                           person_education + person_emp_exp + 
                           person_home_ownership + age_exp_ratio + 
                           credit_hist_per_age,
                         data = train_reg,
                         ntree = 500,
                         mtry = 3,
                         importance = TRUE)

print(rf_model)

```

```{r rf-variable-importance, fig.width=10, fig.height=6}
varImpPlot(rf_model, main = "Random Forest: Variable Importance")
```

```{r rf-error-convergence, fig.width=8, fig.height=5}
plot(rf_model, main = "Random Forest: Error Convergence")

```

## Model Comparison: Regression

We compare all three models on the **test set** using standard regression metrics.

```{r regression-predictions}
# Predictions on test set
# Linear Regression (back-transform from log)
pred_lm   <- exp(predict(lm_step, newdata = test_reg))
# Decision Tree
pred_dt   <- predict(dt_model, newdata = test_reg)
# Random Forest
pred_rf   <- predict(rf_model, newdata = test_reg)

actual <- test_reg$person_income
```

```{r regression-metrics}
# Calculate metrics for each model
calc_metrics <- function(actual, predicted, model_name) {
  data.frame(
    Model     = model_name,
    RMSE      = round(rmse(actual, predicted), 2),
    MAE       = round(mae(actual, predicted), 2),
    R_Squared = round(1 - sum((actual - predicted)^2) / sum((actual - mean(actual))^2), 4),
    MAPE      = round(mean(abs((actual - predicted) / actual)) * 100, 2)
  )
}

comparison_reg <- rbind(
  calc_metrics(actual, pred_lm, "Linear Regression (log)"),
  calc_metrics(actual, pred_dt, "Decision Tree"),
  calc_metrics(actual, pred_rf, "Random Forest")
)

print(comparison_reg)
```

```{r regression-comparison-viz, fig.width=10, fig.height=5}
comparison_long <- comparison_reg %>%
  pivot_longer(cols = c(RMSE, MAE, R_Squared), names_to = "Metric", values_to = "Value")

ggplot(comparison_long, aes(x = Model, y = Value, fill = Model)) +
  geom_col(alpha = 0.8) +
  facet_wrap(~Metric, scales = "free_y") +
  labs(title = "Regression Model Comparison", y = "Value") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1), legend.position = "none")
```

### Actual vs Predicted Scatter Plots

```{r actual-vs-predicted, fig.width=16, fig.height=6}
pred_df <- data.frame(
  Actual = rep(actual, 3),
  Predicted = c(pred_lm, pred_dt, pred_rf),
  Model = rep(c("Linear Regression", "Decision Tree", "Random Forest"),
              each = length(actual))
)

ggplot(pred_df, aes(x = Actual, y = Predicted)) +
  geom_point(alpha = 0.15, size = 1, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed", linewidth = 1) +
  facet_wrap(~Model, ncol = 3) +
  scale_x_continuous(labels = scales::comma, limits = c(0, 300000)) +
  scale_y_continuous(labels = scales::comma, limits = c(0, 300000)) +
  labs(title = "Actual vs Predicted Income",
       subtitle = "Points on the red line = perfect prediction",
       x = "Actual Income (USD)", y = "Predicted Income (USD)") +
  theme(
    plot.title    = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.text.x   = element_text(size = 9, angle = 45, hjust = 1),
    axis.text.y   = element_text(size = 9),
    axis.title    = element_text(size = 13),
    strip.text    = element_text(size = 13, face = "bold")
  )
```

### Final Regression Model Selection

```{r regression-conclusion}
best_model_idx <- which.max(comparison_reg$R_Squared)
cat("BEST REGRESSION MODEL :", "\n")
cat("Model:", comparison_reg$Model[best_model_idx], "\n")
cat("R-squared:", comparison_reg$R_Squared[best_model_idx], "\n")
cat("RMSE:", comparison_reg$RMSE[best_model_idx], "\n")
cat("MAE:", comparison_reg$MAE[best_model_idx], "\n")
```

# Part B: Classification Loan Default Prediction

## Objective

Predict whether a loan applicant will default (`loan_status = 0`) or repay (`loan_status = 1`) using all available features, including engineered ones

## Prepare Classification Data

```{r classification-data-prep}
clf_df <- df %>%
  dplyr::select(loan_status, person_age, person_gender, person_education, person_income,
                person_emp_exp, person_home_ownership, loan_amnt, loan_intent, 
                loan_int_rate, loan_percent_income, cb_person_cred_hist_length,
                credit_score, previous_loan_defaults_on_file,
                loan_credit_ratio, age_exp_ratio, high_risk) %>%
  drop_na()

clf_df$person_education <- factor(clf_df$person_education, ordered = FALSE)

cat("Classification dataset:", nrow(clf_df), "rows x", ncol(clf_df), "columns\n")
```

## Train-Test Split

```{r train-test-split-clf}
set.seed(42)
train_idx_clf <- createDataPartition(clf_df$loan_status, p = 0.8, list = FALSE)
train_clf <- clf_df[train_idx_clf, ]
test_clf  <- clf_df[-train_idx_clf, ]

cat("Training:", nrow(train_clf), "| Test:", nrow(test_clf), "\n")

```

## Model: Logistic Regression (Classification)

Logistic regression is the standard baseline for binary classification problems and is the natural extension of linear regression for binary outcomes.

```{r}
log_model <- glm(loan_status ~ ., data = train_clf, family = binomial(link = "logit"))
log_step  <- stepAIC(log_model, direction = "both", trace = FALSE)

prob_log  <- predict(log_step, newdata = test_clf, type = "response")
pred_log  <- factor(ifelse(prob_log > 0.5, 1, 0), levels = c("0", "1"))

cat("\n Logistic Regression \n")
cm_log <- confusionMatrix(pred_log, test_clf$loan_status, positive = "1")
print(cm_log)
```

## Model: Random Forest Classifier

```{r random-forest-classifier}
set.seed(42)
rf_clf <- randomForest(loan_status ~ ., data = train_clf, 
                       ntree = 500, mtry = 4, importance = TRUE)

pred_rf_clf <- predict(rf_clf, newdata = test_clf)
prob_rf_clf <- predict(rf_clf, newdata = test_clf, type = "prob")[, 2]

cat("\n Random Forest \n")
cm_rf <- confusionMatrix(pred_rf_clf, test_clf$loan_status, positive = "1")
print(cm_rf)

```

## Model: Decision Tree Classifier

```{r decision-tree-clf}
dt_clf <- rpart(loan_status ~ ., data = train_clf, method = "class",
                control = rpart.control(cp = 0.001, maxdepth = 8))

pred_dt_clf <- predict(dt_clf, newdata = test_clf, type = "class")
prob_dt_clf <- predict(dt_clf, newdata = test_clf, type = "prob")[, 2]

cat("\n=== Decision Tree ===\n")
cm_dt <- confusionMatrix(pred_dt_clf, test_clf$loan_status, positive = "1")
print(cm_dt)
```

```{r dt-clf-viz, fig.width=14, fig.height=8}
rpart.plot(dt_clf, type = 4, extra = 104, roundint = FALSE,
           main = "Decision Tree: Loan Default Classification",
           box.palette = "RdGn", shadow.col = "gray")

```

## Classification Model Comparison

### Confusion Matrices

```{r confusion-matrices}
cat("Logistic Regression\n")
cm_log <- confusionMatrix(pred_log, test_clf$loan_status, positive = "1")
print(cm_log)

cat("\n Random Forest \n")
cm_rf <- confusionMatrix(pred_rf_clf, test_clf$loan_status, positive = "1")
print(cm_rf)

cat("\n Decision Tree \n")
cm_dt <- confusionMatrix(pred_dt_clf, test_clf$loan_status, positive = "1")
print(cm_dt)
```

### ROC Curves and AUC

**ROC** stands for **Receiver Operating Characteristic**. It's a curve that plots two things as you change the classification threshold (the 0.5 cutoff).\
**AUC** stands for **Area Under the Curve**. It's a single number that summarizes the entire ROC curve:

-   **AUC = 1.0** perfect model, never makes mistakes

-   **AUC = 0.9–1.0** excellent model

-   **AUC = 0.8–0.9** good model

-   

    ### Credit Risk Analysis

    **AUC = 0.7–0.8** fair model

    **Why we use AUC instead of accuracy for this project:** Your dataset is imbalanced — 78% defaults, 22% repaid. A model that blindly predicts "default" for everyone would get 78% accuracy, which looks impressive but is actually useless. AUC is not fooled by this trick because it evaluates how well the model **separates** the two classes, regardless of their proportions.

```{r roc-curves, fig.width=8, fig.height=6}
roc_log <- roc(test_clf$loan_status, prob_log, levels = c("0", "1"), 
               direction = "<")
roc_rf  <- roc(test_clf$loan_status, prob_rf_clf, levels = c("0", "1"), 
               direction = "<")
roc_dt  <- roc(test_clf$loan_status, prob_dt_clf, levels = c("0", "1"), 
               direction = "<")

plot(roc_log, col = "#2980b9", lwd = 2, main = "ROC Curves: Classification Models")
plot(roc_rf, col = "#27ae60", lwd = 2, add = TRUE)
plot(roc_dt, col = "#e74c3c", lwd = 2, add = TRUE)
abline(a = 0, b = 1, lty = 2, col = "gray50")

legend("bottomright", 
       legend = c(
         paste("Logistic Regression (AUC =", round(auc(roc_log), 4), ")"),
         paste("Random Forest (AUC =", round(auc(roc_rf), 4), ")"),
         paste("Decision Tree (AUC =", round(auc(roc_dt), 4), ")")
       ),
       col = c("#2980b9", "#27ae60", "#e74c3c"), lwd = 2, cex = 0.9)
```

### Summary Comparison Table

```{r classification-summary}
clf_comparison <- data.frame(
  Model = c("Logistic Regression", "Random Forest", "Decision Tree"),
  Accuracy = c(cm_log$overall["Accuracy"], cm_rf$overall["Accuracy"], 
               cm_dt$overall["Accuracy"]),
  Sensitivity = c(cm_log$byClass["Sensitivity"], cm_rf$byClass["Sensitivity"],
                  cm_dt$byClass["Sensitivity"]),
  Specificity = c(cm_log$byClass["Specificity"], cm_rf$byClass["Specificity"],
                  cm_dt$byClass["Specificity"]),
  F1_Score = c(cm_log$byClass["F1"], cm_rf$byClass["F1"], cm_dt$byClass["F1"]),
  AUC = c(auc(roc_log), auc(roc_rf), auc(roc_dt))
)

clf_comparison <- clf_comparison %>% 
  mutate(across(where(is.numeric), ~round(., 4)))

print(clf_comparison)
```

```{r classification-viz, fig.width=12, fig.height=5}
clf_long <- clf_comparison %>%
  pivot_longer(cols = -Model, names_to = "Metric", values_to = "Value")

ggplot(clf_long, aes(x = Model, y = Value, fill = Model)) +
  geom_col(alpha = 0.8) +
  facet_wrap(~Metric, scales = "free_y", ncol = 5) +
  labs(title = "Classification Model Performance Comparison", y = "Score") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1), legend.position = "none") +
  scale_fill_manual(values = c("#2980b9", "#e74c3c", "#27ae60"))
```

### Best Classification Model

```{r classification-conclusion}
best_clf_idx <- which.max(clf_comparison$AUC)
cat("BEST CLASSIFICATION MODEL","\n")

cat("Model:", clf_comparison$Model[best_clf_idx], "\n")
cat("AUC:", clf_comparison$AUC[best_clf_idx], "\n")
cat("Accuracy:", clf_comparison$Accuracy[best_clf_idx], "\n")
cat("F1 Score:", clf_comparison$F1_Score[best_clf_idx], "\n")
```

## Summary of Findings

```{r final-regression-summary}
cat("Best Regression Model:", comparison_reg$Model[best_model_idx], "\n")
cat("Test R²:", comparison_reg$R_Squared[best_model_idx], "\n")
cat("Test RMSE:", comparison_reg$RMSE[best_model_idx], "\n\n")

cat("Key regression findings:\n")

cat("- Log transformation of income was necessary to meet normality assumptions.\n")
cat("- Stepwise selection (direction='both') was used to identify optimal predictors.\n")
cat("- All four OLS assumptions were checked: linearity, normality, ",
    "homoscedasticity, multicollinearity.\n")
```

### Classification (Loan Default Prediction)

```{r final-classification-summary}

cat("Best Classification Model:", clf_comparison$Model[best_clf_idx], "\n")
cat("AUC:", clf_comparison$AUC[best_clf_idx], "\n")
cat("Accuracy:", clf_comparison$Accuracy[best_clf_idx], "\n\n")

cat("Key classification findings:\n")

cat("- The target variable is imbalanced (78% default, 22% repaid).\n")
cat("- ROC-AUC is the primary evaluation metric due to class imbalance.\n")
cat("- Feature engineering (loan_credit_ratio, high_risk flag) added ",
    "predictive value.\n")

```

### Credit Risk Analysis

```{r credit-risk-summary}
cat("Credit Risk Insights:\n")
cat("- Higher credit scores are associated with lower default rates.\n")
cat("- loan_percent_income is the strongest predictor of default.\n")
cat("- Previous loan defaults strongly predict future default behavior.\n")
cat("- High-risk applicants (high loan%, low credit) default at much higher rates.\n")

```

## Limitations

-   The dataset has no missing values, which is unrealistic for real-world data.
-   Income prediction using only personal demographics has inherent limitations occupation and industry data would significantly improve the model.
-   The class imbalance in loan_status (78/22) was not addressed with resampling techniques.
-   Cross-validation was not performed here but would strengthen the reliability of all results.
